### 2022-11-01 Papers/ICCC-2023/sections/design.tex
even more corrections

CTFEF apporach was created during implementation of the first compiler for the C-=-1 language\cite{grabski2022compilation}.
CTFEF compiler has four major components: Frontend, Interpreter, Compiler Interface and Backend.
Figure \ref{CTFE-first-compiler-structure} contains a diagram with an overview of how these parts interact with eachother, during the compilaton process.
Frontend, described in chapter \ref{frontend}, parses the code in the compiled language and constructs its intermediate representation, using interpreter's data structures.
It is used to analyze both user code and the Compiler Interface.
After the intermediate representation is constructed, it is passed to the Interpreter, which was described in chapter \ref{interpreter}.
Compiler Interface intermediate representation is then executed, using the user program as data.
This step converts the semantic model of the program into the backends intermediate language.
This process is further explained in chapter \ref{compiler-interface}.
Finaly the Backend generates the executable file.

@@

\subsection{Frontend}
\label{frontend}

In the CTFEF approach, frontend serves the same role of constructing the programs intermediate representation, as in conventional compilers \cite{puntambekar:compiler_design}.
The major difference lays in the data structures used to describe the program.
For a CTFEF compiler to work, they must be accessible to the program running within the interpreter.
The increace of complexity, caused by using data structures accessible from interpreted code, depends on the design of the interpreter.

@@

In CTFEF, interpreter is main component of the compiler.
It executes the Compiler-Interface which translates the intermediate representation into the backend's assembly and serves as what is sometimes called the 'middle-end' of the compiler\cite{hsu2021llvm}.
To do it, it must be able to treat the program's intermediate representation both as code and data.

@@

Compiler Interface translates the program's intermediate representation into the backend's assembly language.
This component is interpreted during compilation.
What is unique about CTFEF is that this part of the compiler can thus be written in the target language, during Stage 0 of the compiler bootstrapping process, as in the case of the C-=-1 compiler\cite{puntambekar:compiler_design, novillo2007gcc, grabski2022compilation}.

Compiler Interface is a regular code package that contains a function marked as the Compiler Interface Entry-point.
That procedure must accept a set of modules to be compiled and a Compilation Context that is used to generate the Backends assembly.
The module descriptors that are passed to the Compiler Interface are built by the frontend, as can be seen in figure \ref{CTFE-first-compiler-structure}.

After the Compiler Interface has finished generating backend assembly, the Compiler Backend is invoked to generate the binary executable.
\subsection{Backend}
\label{backend}

CTFEF does not put any additional requirements on compiler backend.
When using this pattern, an out-of-the-box backend library, can be used, as was the case with C-=-1 compiler.

The backend code must be invokable from within the interpreted program in the target language.
Depending on how the interpreter was designed, this may require significant effort. %todo: maybe other word
Compiler backends are large and for the Compiler Interface to take advantage of them, they must be fully available.
This means exposing each function and type within the library, to the interpreted code.
These bindings could feasibly be generated automatically \cite{marshalling_auto_generation}, but this technique was not used when implementing C-=-1 compiler.

>>>>

CTFEF apporach was created during implementation of the first compiler for the C-=-1 language\cite{grabski2022compilation}.
CTFEF compiler has four major components: 
\begin{enumerate}
	\item Frontend
	\item Interpreter
	\item Compiler Interface
	\item Backend
\end{enumerate}

Figure \ref{CTFE-first-compiler-structure} contains a diagram with an overview of how these parts interact with eachother, during the compilaton process.
Frontend, described in chapter \ref{Frontend}, parses the code in the compiled language and constructs its intermediate representation, using interpreter's data structures.
It is used to analyze both user code and the Compiler Interface.
After the intermediate representation is constructed, it is passed to the Interpreter, which was described in chapter \ref{interpreter}.
Compiler Interface intermediate representation is then executed, using the user program as data.
This step converts the semantic model of the program into the Backends intermediate language.
This process is further explained in chapter \ref{compiler-interface}.
Finaly the Backend generates the executable file.

@@

\subsection{Frontend}
\label{Frontend}

In the CTFEF approach, Frontend serves the same role of constructing the programs intermediate representation, as in conventional compilers \cite{puntambekar:compiler_design}.
The major difference lays in the data structures used to describe the program.
For a CTFEF compiler to work, they must be accessible to the program running within the interpreter.
This may make frontend more complex.
The additional challenge of representing a user program, using interpreter data structures depends on the design of the interpreter.

@@

In CTFEF, interpreter is main component of the compiler.
It executes the Compiler-Interface which translates the intermediate representation into the Backend's assembly and serves as what is sometimes called the 'middle-end' of the compiler\cite{hsu2021llvm}.
To do it, it must be able to treat the program's intermediate representation both as code and data.

@@

Compiler Interface translates the program's intermediate representation into the Backend's assembly language.
This component is interpreted during compilation.
What is unique about CTFEF is that this part of the compiler can thus be written in the target language, during Stage 0 of the compiler bootstrapping process, as in the case of the C-=-1 compiler\cite{puntambekar:compiler_design, novillo2007gcc, grabski2022compilation}.

Compiler Interface is a regular code package that contains a function marked as the Compiler Interface Entry-point.
That procedure must accept a set of modules to be compiled and a Compilation Context that is used to generate the Backends assembly.
The module descriptors that are passed to the Compiler Interface are built by the Frontend, as can be seen in figure \ref{CTFE-first-compiler-structure}.

After the Compiler Interface has finished generating Backend assembly, the Compiler Backend is invoked to generate the binary executable.
\subsection{Backend}
\label{Backend}

CTFEF does not put any additional requirements on compiler Backend.
When using this pattern, a generic Backend library, can be used.
C-=-1 compiler used LLVM\cite{lattner2008llvm} as its backend.

The Backend code must be invokable from within the interpreted program in the target language.
Depending on how the interpreter was designed, this may require significant effort. %todo: maybe other word
Compiler Backends are large and for the Compiler Interface to take advantage of them, their entire interface must be fully available in the interpreted context.
This means exposing each function and type within the library, to the interpreted code, by duplicating their signatures.
These bindings could feasibly be generated automatically \cite{marshalling_auto_generation}, but this technique was not used when implementing C-=-1 compiler.

### 2022-11-01 Papers/ICCC-2023/sections/implementation.tex
even more corrections

The first CTFEF was created for the C-=-1 language, using out-of-the-box parser generator and backend.
The most important aspect of implementing a CTFEF compiler is the design of the data structures, described in chapter \ref{data_structures}, that will be used by the interpreter.

@@

The final part of a CTFEF compiler is the Backend Interface.
It is a program, written in the target language, and ran at compile-time that translates the semantic model into the backends' assembly language.
Backend Interface implemented for C-=-1 is relativley small and is described in chapter \ref{implementation/backend-interface}.

@@

To aid in generating meaningfull error messages, all elements of the semantic model, have a \lstinline{sourcePointers}.
It is a simple structure, that contains the filename and the line number of the expression or instruction.
This information can be passed to compiler intrinsic functions, such as \lstinline{raiseError}, to generate error messages for the user.
Listing \ref{lst:noDiscardCm1} contains an example use of this functionality.

@@

\subsection{Backend interface}
\label{implementation/backend-interface}

The C-=-1 backend interface uses LLVM \cite{llvmir} code generation API that has been exposed by the compiler.
The functionality which has been made available to C-=-1 represents a minimal subset of LLVM, that is sufficent to implement a basic compiler.

@@

One of the possible ways of achieving this, is to declare an interface for an attribute generating a functions symbol name.
Listing \ref{lst:selected_backend_interface} contains relevant code of a Backend Interface that uses such an attribute, to override mark external symbols.
Interface \lstinline{ISymbolNameOverride} contains only one method: \lstinline{createSymbolName} that returns the name of the symbol in the generated assembly.

Functions \lstinline{buildFunction} and \lstinline{getFunctionName}, from Listing \ref{lst:selected_backend_interface}, are parts of Backend Interface.
They are invoked in order to convert a C-=-1 \lstinline{functionDescriptor} to a LLVMIR function.
They were included in the Listing, because they are the only parts of the Backend Intreface that need to interact with \lstinline{ISymbolNameOverride} attributes.
Both of these procedures, check whether an attribute implementing this interface is attached to the function they are currently processing.
This happens on lines three and nineteen of Listing \ref{lst:selected_backend_interface}.
If that attribute is present, code of that function is ignored, and it is treated as an external symbol: condition on line twenty ommits execution of \lstinline{build_block} function on line twenty-three.
Procedure \lstinline{getFunctionName} contains a similar condition on line three, that decides how the name should be generated.
If an \lstinline{ISymbolNameOverride} attribute is present, it will be created by \lstinline{createSymbolName} method of the attribute attached to the function.
Otherwise, the \lstinline{mangleName} function will create name of function's symbol based on its parameters and return type.

@@

	\begin{lstlisting}[
	  numbers=left,
	  firstnumber=0,
	  caption={Part of a Backend Interface, using \lstinline{ISymbolNameOverride} interface},
	  aboveskip=0pt,
	  label={lst:selected_backend_interface}
	  ]

>>>>

The first CTFEF was created for the C-=-1 language, using generic parser generator and Backend.
The most important aspect of implementing a CTFEF compiler is the design of the data structures, described in chapter \ref{data_structures}, that will be used by the interpreter.

@@

The final part of a CTFEF compiler is the Backend Interface.
It is a program, written in the target language, and ran at compile-time that translates the semantic model into the Backends' assembly language.
Backend Interface implemented for C-=-1 is relativley small and is described in chapter \ref{implementation/Backend-interface}.

@@

All elements of the semantic model, have a \lstinline{sourcePointers}.
It is a simple structure, that contains the filename and the line number of the expression or instruction.
This information can be passed to compiler intrinsic functions, such as \lstinline{raiseError}, to generate error messages for the user.
Listing \ref{lst:noDiscardCm1} contains an example use of this functionality.
The \lstinline{pointerToSource} makes the messages generated by the compiler easier to understand for the programmer.

@@

\subsection{Backend interface}
\label{implementation/Backend-interface}

The C-=-1 Backend interface uses LLVM \cite{llvmir} code generation API that has been exposed by the compiler.
The functionality which has been made available to C-=-1 represents a minimal subset of LLVM, that is sufficent to implement a basic compiler.

@@

One of the possible ways of achieving this, is to declare an interface for an attribute generating a functions symbol name.
Listing \ref{lst:selected_Backend_interface} contains relevant code of a Backend Interface that uses such an attribute, to override mark external symbols.
Interface \lstinline{ISymbolNameOverride} contains only one method: \lstinline{createSymbolName} that returns the name of the symbol in the generated assembly.

Functions \lstinline{buildFunction} and \lstinline{getFunctionName}, from Listing \ref{lst:selected_Backend_interface}, are parts of Backend Interface.
They are invoked in order to convert a C-=-1 \lstinline{functionDescriptor} to a LLVMIR function.
They were included in the Listing, because they are the only parts of the Backend Intreface that need to interact with \lstinline{ISymbolNameOverride} attributes.
Both of these procedures, check whether an attribute implementing this interface is attached to the function they are currently processing.
This happens on lines three and nineteen of Listing \ref{lst:selected_Backend_interface}.
If that attribute is present, code of that function is ignored, and it is treated as an external symbol: condition on line twenty ommits execution of \lstinline{build_block} function on line twenty-three.
Procedure \lstinline{getFunctionName} contains a similar condition on line three, that decides how the name should be generated.
If an \lstinline{ISymbolNameOverride} attribute is present, it will be created by \lstinline{createSymbolName} method of the attribute attached to the function.
Otherwise, the \lstinline{mangleName} function will create name of function's symbol based on its parameters and return type.

@@

	\begin{lstlisting}[
	  numbers=left,
	  firstnumber=0,
	  caption={Part of a Backend Interface, using \lstinline{ISymbolNameOverride} interface},
	  aboveskip=0pt,
	  label={lst:selected_Backend_interface}
	  ]
